Combinatorial Proofs and Decomposition Theorems for First-order Logic

### 2021-01-16 main.tex
still in IX

\newcommand{\graph}[1]{\mathcal{#1}}
\newcommand{\vertices}[1][]{\ifthenelse{\isempty{#1}}{V}{V_{\graph{#1}}}}
\newcommand{\edges}[1][]{\ifthenelse{\isempty{#1}}{E}{E_{\graph{#1}}}}
\newcommand{\bgraph}[1]{\mathcal{\vec{#1}}}
\newcommand{\lgraph}[1]{\mathcal{#1}^{\mathsf{L}}}
\newcommand{\leaps}[1]{L_{#1}}

\newcommand{\vgraphof}[1]{V_{\graphof{#1}}}
\newcommand{\gA}{\graph{A}}
\newcommand{\gB}{\graph{B}}
\newcommand{\gC}{\graph{C}}
\newcommand{\gG}{\graph{G}}
\newcommand{\gH}{\graph{H}}
\newcommand{\gN}{\graph{N}}

\newcommand{\bA}{\bgraph{A}}
\newcommand{\bC}{\bgraph{C}}
\newcommand{\bG}{\bgraph{G}}
\newcommand{\bH}{\bgraph{H}}

\newcommand{\vA}{\vertices[A]}
\newcommand{\vC}{\vertices[C]}
\newcommand{\vG}{\vertices[G]}
\newcommand{\vH}{\vertices[H]}

@@

\newcommand\aiD {\mathsf{ai}}
\newcommand\faD {\forall}
\newcommand\exD {\exists}
\newcommand\tttD {\ttt}
\renewcommand\wD {\mathsf{w}}
\newcommand\wlD {\mathsf{w_l}}
\newcommand\wrD {\mathsf{w_r}}
\renewcommand\cD {\mathsf{c}}
\renewcommand\acD {\mathsf{ac}}
\newcommand\mfaD {\mathsf{m_\forall}}
\newcommand\mexD {\mathsf{m_\exists}}

@@

\todo{this has to be moved to the right place at some point}

The \bfit{propositional encoding} $\PE{A}$ of a formula $A$ is defined inductively by:
\begin{equation*}
  \begin{array}{r@{\;=\;}l}
    \PE{a} & a\\
    \PE{(A \cor B)} &  \PE{A} \cor \PE{B} \\
    \PE{(A \cand B)} & \PE{A} \cand \PE{B}
  \end{array}
  \qquad
  \begin{array}{r@{\;=\;}l}
    \PE{(\forall x A)} & u_x \cor \PE{A}\\
    \PE{(\exists x A)} & e_x \cand \PE{A}
  \end{array}
\end{equation*}
where $u_x$ and $e_x$ are fresh nullary atoms. The purpose of the propositional encoding is to simplify some arguments that do not depend on the variable bindings. In particular, observe that $\graphof A$ and $\graphof{\PE A}$ only differ in the labelling of certain vertices.

@@

@@

If a derivation $\Deri$ contains instances of the rules $\mfaD$ and
$\mexD$ we can no longer naively define the rectification
$\rectif\Deri$ as in the previous section for $\FOMLS$, as these two
rules cannot be applied if premise and conclusion are rectified. For
this reason we define here rectified versions $\rectif\mfaD$ and
$\rectif\mexD$, show below:
\begin{equation*}
  \vlinf{\rectif\mfaD}{}{\Scons{\forall x.(Ax\vlor Bx)}}{\Scons{(\forall y.Ay)\vlor(\forall z.Bz)}}
  \qquad
  \vlinf{\rectif\mexD}{}{\Scons{\exists x.(Ax\vlor Bx)}}{\Scons{(\exists y.Ay)\vlor(\exists z.Bz)}}
\end{equation*}
Here, we use the notation $A\cdot$ for a formula $A$ with occurrences
of a placeholder $\cdot$ for a variable. Then $Ax$ stands for the
results of replacing that placeholder with $x$, and also indicating
that $x$ must not occur in $A\cdot$. Then $\forall x.Ax$ and $\forall
y.Ay$ are the same formula modulo renaming of the bound variable bound
by the outermost $\forall$-quantifier. In the rules $\rectif\mfaD$ and
$\rectif\mexD$ we also demand that the variables $x$, $y$, and $z$ do
not occur in the context $\Sconhole$.

For each instance $\vlinf{\rr}{}{P}{Q}$ of an inference rule in
$\set{\wlD,\wrD,\acD,\me,\rectif\mfaD,\rectif\mexD,\fequ}$ we can
define the \bfit{induced map} $\mapof{\rr}\colon\vgraphof
Q\to\vgraphof P$ which acts as the identity for
$\rr\in\set{\me,\fequ}$ and as the canonical injection for
$\rr\in\set{\wlD,\wrD}$. For $\rr=\acD$ it maps the vertices
corresponding to the two atoms in the premise to vertex of the
contracted atom in the conclusion, and for
$\rr\in\set{\rectif\mfaD,\rectif\mexD}$ it maps the two vertices
corresponding to the quantifiers in the premise to the one in the
conclusion (as acts as the identity on all other vertices). For a
derivation $\Deri$ in
$\set{\wlD,\wrD,\acD,\me,\rectif\mfaD,\rectif\mexD,\fequ}$ we define
the \bfit{induced map} $\mapof\Deri$ as the composition of the induced
maps of the rule instances in $\Deri$.

\todo{example}
\begin{lemma}
  Let $\vlderivation{
    \vlde{\set{\wlD,\wrD,\acD,\me,\rectif\mfaD,\rectif\mexD,\fequ}}{\Deri}{B}{
      \vlhy{A}}}$ be a rectified derivation. Then the induced map
  $\mapof\Deri\colon\graphof{A}\to\graphof{B}$ is a skew bifibration.
\end{lemma}

\begin{proof}
  For $\Deri$, we construct its propositional encoding $\PE\Deri$ by
  translating every line into its proposional encoding. The rules
  $\rectif\mfaD$ and $\rectif\mexD$ are replaced by:
  \begin{equation*}
    \vlderivation{
      \vlin{\rectif\acD}{}{\Scons{u_x\vlor\PE{(Ax)}\vlor\PE{(Ax)}}}{
        \vlin{\fequ}{}{\Scons{(u_y\vlor u_z)\vlor(\PE{(Ax)}\vlor\PE{(Ax)})}}{
          \vlhy{\Scons{(u_y\vlor\PE{(Ax)})\vlor (u_z\vlor\PE{(Ax)})}}}}}
  \end{equation*}
  and
  \begin{equation*}
    \vlderivation{
      \vlin{\rectif\acD}{}{\Scons{e_x\vlan\PE{(Ax)}\vlor\PE{(Ax)}}}{
        \vlin{\me}{}{\Scons{(e_y\vlor e_z)\vlan(\PE{(Ax)}\vlor\PE{(Ax)})}}{
          \vlhy{\Scons{(e_y\vlan\PE{(Ax)})\vlor (e_z\vlan\PE{(Ax)})}}}}}
  \end{equation*}
  respectively. This gives us a derivation $\vlderivation{
    \vlde{\set{\wlD,\wrD,\acD,\rectif\acD,\me,\fequ}}{\PE\Deri}{\PE
      B}{ \vlhy{\PE A}}}$ with $\mapof{\PE\Deri}$ being equal to
  $\mapof\Deri$, modulo relabeling of of the binders. It has been
  shown in~\cite{str:07:RTA} that $\mapof{\PE\Deri}$ is a skew
  fibration (see
  also~\cite{hughes:invar,str:ral:tableaux19,str:RR-9048}). Hence,
  $\mapof\Deri$ is a skew fibration. \todo{show the other properties,
    once they are fixed}
\end{proof}
\subsection{From Skew Bifibrations to Contraction and Weakening}

\begin{lemma}
  Let $\phi\colon\gA\to\gB$ be a skew bifibration, and let $A$ and $B$
  be formulas with $\graphof A=\gA$ and $\graphof B=\gB$. Then there
  is a derivation $\vlderivation{
    \vlde{\set{\wlD,\wrD,\acD,\me,\rectif\mfaD,\rectif\mexD,\fequ}}{\Deri}{B}{
      \vlhy{A}}}$ with $\mapof\Deri=\phi$.
\end{lemma}

\begin{proof}
  Let $\PE\gA=\graphof{\PE A}$ and $\PE\gB=\graphof{\PE B}$ which are equal to $\gA$ and $\gB$, respectively, modulo relabeling of the binders. Thus, we can define $\PE\phi\colon\PE\gA\to\PE\gB$, which is a skew fibration as well. 
\end{proof}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

>>>>

\newcommand{\graph}[1]{\mathcal{#1}}
\newcommand{\vertices}[1][]{\ifthenelse{\isempty{#1}}{V}{V_{{\graph{#1}}}}}
\newcommand{\edges}[1][]{\ifthenelse{\isempty{#1}}{E}{E_{\graph{#1}}}}
\newcommand{\bgraph}[1]{\mathcal{\vec{#1}}}
\newcommand{\lgraph}[1]{\mathcal{#1}^{\mathsf{L}}}
\newcommand{\leaps}[1]{L_{#1}}

\newcommand{\vgraphof}[1]{V_{\graphof{#1}}}
\newcommand{\gA}{\graph{A}}
\newcommand{\gB}{\graph{B}}
\newcommand{\gBone}{\graph{B}_1}
\newcommand{\gC}{\graph{C}}
\newcommand{\gG}{\graph{G}}
\newcommand{\gH}{\graph{H}}
\newcommand{\gN}{\graph{N}}

\newcommand{\bA}{\bgraph{A}}
\newcommand{\bB}{\bgraph{B}}
\newcommand{\bC}{\bgraph{C}}
\newcommand{\bG}{\bgraph{G}}
\newcommand{\bH}{\bgraph{H}}

\newcommand{\vA}{\vertices[A]}
\newcommand{\vB}{\vertices[B]}
\newcommand{\vBone}{V_{\mathcal{B}_1}}
\newcommand{\vC}{\vertices[C]}
\newcommand{\vG}{\vertices[G]}
\newcommand{\vH}{\vertices[H]}

@@

\newcommand\aiD {\mathsf{ai}}
\newcommand\faD {\forall}
\newcommand\exD {\exists}
\newcommand\tttD {\ttt}
\renewcommand\wD {\mathsf{w}}
\newcommand\wlD {\mathsf{w_l}}
\newcommand\wrD {\mathsf{w_r}}
\renewcommand\cD {\mathsf{c}}
\renewcommand\acD {\mathsf{ac}}
\newcommand\acDeq {\mathsf{ac}^\fequ}
\newcommand\mfaD {\mathsf{m_\forall}}
\newcommand\mexD {\mathsf{m_\exists}}

@@

@@

In this section we establish the relation between ske bifibrations and derivations  in
$\set{\wlD,\wrD,\acD,\me,\mfaD,\mexD,\fequ}$. However, 
if a derivation $\Deri$ contains instances of the rules $\mfaD$ and
$\mexD$ we can no longer naively define the rectification
$\rectif\Deri$ as in the previous section for $\FOMLS$, as these two
rules cannot be applied if premise and conclusion are rectified. For
this reason we define here rectified versions $\rectif\mfaD$ and
$\rectif\mexD$, show below:
\begin{equation*}
  \vlinf{\rectif\mfaD}{}{\Scons{\forall x.(Ax\vlor Bx)}}{\Scons{(\forall y.Ay)\vlor(\forall z.Bz)}}
  \qquad
  \vlinf{\rectif\mexD}{}{\Scons{\exists x.(Ax\vlor Bx)}}{\Scons{(\exists y.Ay)\vlor(\exists z.Bz)}}
\end{equation*}
Here, we use the notation $A\cdot$ for a formula $A$ with occurrences
of a placeholder $\cdot$ for a variable. Then $Ax$ stands for the
results of replacing that placeholder with $x$, and also indicating
that $x$ must not occur in $A\cdot$. Then $\forall x.Ax$ and $\forall
y.Ay$ are the same formula modulo renaming of the bound variable bound
by the outermost $\forall$-quantifier. In the rules $\rectif\mfaD$ and
$\rectif\mexD$ we also demand that the variables $x$, $y$, and $z$ do
not occur in the context $\Sconhole$.

For each instance $\vlinf{\rr}{}{P}{Q}$ of an inference rule in
$\set{\wlD,\wrD,\acD,\me,\rectif\mfaD,\rectif\mexD,\fequ}$ we can
define the \bfit{induced map} $\mapof{\rr}\colon\vgraphof
Q\to\vgraphof P$ which acts as the identity for
$\rr\in\set{\me,\fequ}$ and as the canonical injection for
$\rr\in\set{\wlD,\wrD}$. For $\rr=\acD$ it maps the vertices
corresponding to the two atoms in the premise to vertex of the
contracted atom in the conclusion, and for
$\rr\in\set{\rectif\mfaD,\rectif\mexD}$ it maps the two vertices
corresponding to the quantifiers in the premise to the one in the
conclusion (as acts as the identity on all other vertices). For a
derivation $\Deri$ in
$\set{\wlD,\wrD,\acD,\me,\rectif\mfaD,\rectif\mexD,\fequ}$ we can then define
the \bfit{induced map} $\mapof\Deri$ as the composition of the induced
maps of the rule instances in $\Deri$.

\todo{example}

@@

\begin{lemma}\label{lem:cw->skew}
  Let $\vlderivation{
    \vlde{\set{\wlD,\wrD,\acD,\me,\rectif\mfaD,\rectif\mexD,\fequ}}{\Deri}{B}{
      \vlhy{A}}}$ be a rectified derivation. Then the induced map
  $\mapof\Deri\colon\graphof{A}\to\graphof{B}$ is a skew bifibration.
\end{lemma}

Before we show the proof of this lemma, we introduce another useful
concept: the \bfit{propositional encoding} $\PE{A}$ of a formula $A$,
which is a proposional formula with the property that $\graphof{\PE
  A}=\graphof{A}$. For this, we introduce new propositional variables
that have the same names as the (first-order) variables
$x\in\VAR$. Then $\PE{A}$ is defined inductively by:
\begin{equation*}
  \begin{array}{r@{\;=\;}l}
    \PE{a} & a\\
    \PE{(A \cor B)} &  \PE{A} \cor \PE{B} \\
    \PE{(A \cand B)} & \PE{A} \cand \PE{B}
  \end{array}
  \qquad
  \begin{array}{r@{\;=\;}l}
    \PE{(\forall x A)} & x \cor \PE{A}\\
    \PE{(\exists x A)} & x \cand \PE{A}
  \end{array}
\end{equation*}

\begin{lemma}
  \label{lem:PE}
  For every formula $A$, we have $\graphof{\PE A}=\graphof{A}$.
\end{lemma}

\begin{proof}
  Straightforward induction on $A$. 
\end{proof}

We use $\PE\fequ$ to denote the restriction of $\fequ$ to
propositional formulas, i.e., the first two lines in~\eqref{eq:fequ}.

\begin{proof}[Proof of Lemma~\ref{lem:cw->skew}]
  For $\Deri$, we construct its propositional encoding $\PE\Deri$ by
  translating every line into its proposional encoding. The instances of the rules
  $\rectif\mfaD$ and $\rectif\mexD$ are replaced two steps by:
  \begin{equation*}
    \vlderivation{
      \vlin{\rectif\acD}{}{\Scons{x\vlor(\PE{(Ax)}\vlor\PE{(Bx))}}}{
        \vlin{\fequ}{}{\Scons{(y\vlor z)\vlor(\PE{(Ay)}\vlor\PE{(Bz)})}}{
          \vlhy{\Scons{(y\vlor\PE{(Ay)})\vlor (z\vlor\PE{(Bz)})}}}}}
  \end{equation*}
  and
  \begin{equation*}
    \vlderivation{
      \vlin{\rectif\acD}{}{\Scons{x\vlan\PE{(Ax)}\vlor\PE{(Bx)}}}{
        \vlin{\me}{}{\Scons{(y\vlor z)\vlan(\PE{(Ay)}\vlor\PE{(Bz)})}}{
          \vlhy{\Scons{(y\vlan\PE{(Ay)})\vlor (z\vlan\PE{(Bz)})}}}}}
  \end{equation*}
  respectively, where $\rectif\acD$ is a $\acD$ that renames the
  variabes---the atom variable, as well as the first-order variable of
  the same name---as everything is recified, there is no ambiguity
  here.
  This gives us a derivation $\vlderivation{
    \vlde{\set{\wlD,\wrD,\acD,\rectif\acD,\me,\PE\fequ}}{\PE\Deri}{\PE
      B}{ \vlhy{\PE A}}}$ such that  $\mapof{\PE\Deri}=\mapof\Deri$. It has been
  shown in~\cite{str:07:RTA} that $\mapof{\PE\Deri}$ is a skew
  fibration (see
  also~\cite{hughes:invar,str:ral:tableaux19,str:RR-9048}). Hence,
  $\mapof\Deri$ is a skew fibration. \todo{show the other properties,
    once they are fixed}
\end{proof}
\subsection{From Skew Bifibrations to Contraction and Weakening}

\begin{lemma}\label{skew->cw}
  Let $\phi\colon\gA\to\gB$ be a skew bifibration, and let $A$ and $B$
  be formulas with $\graphof A=\gA$ and $\graphof B=\gB$. Then there
  is a derivation $\vlderivation{
    \vlde{\set{\wlD,\wrD,\acD,\me,\rectif\mfaD,\rectif\mexD,\fequ}}{\Deri}{B}{
      \vlhy{A}}}$ with $\mapof\Deri=\phi$.
\end{lemma}

\begin{proof}
  By Lemma~\ref{lem:PE} we have $\gA=\graphof{\PE A}$ and
  $\gB=\graphof{\PE B}$.  Let $\vB'\subseteq\vB$ be the image of
  $\phi$, and let $\gB_1$ be the subgraph of $\gB$ induced by
  $\vB'$. Now let $\PE B_1$ be the propositional formula obtained from
  $\PE B$ by removing all atoms that are not represented by vertices
  in $\vB'$. Then $\graphof{\PE B_1}=\gB_1$. Furthermore, the map
  $\phi'\colon \gB_1\to\gB$ is an injection that reflects
  edges. Hence, by \cite[Proposition~7.6.1]{str:07:RTA}, we
  have a derivation
  $\vlder{\set{\wlD,\wrD}}{\PE\Deri_1}{\PE B}{\PE B_1}$. \todo{show that we have $B_1$ with $\graphof{B_1}=\gB_1$, so that we can can get
  first order derivation $\vlder{\set{\wlD,\wrD}}{\Deri_1}{B}{B_1}$.}

  Now, let $\sigma$ be the substitution induced by $\phi$, and let
  $\gA_1=\gA\sigma$ be the graph obtained from $\gA$ by applying
  $\sigma$ to all the labels. Note that $\gA_1$ is not neccesarily a
  fograph, as binders might be illegal. But it still is a cograph, and
  we have a surjective skew fibration $\phi''\colon\gA_1\to\gB_1$ that
  preserves the labels. Therefore, by
  \cite[Proposition~7.5]{str:ral:tableaux19}, there is a derivation
  $\vlder{\set{\acD,\me,\PE\fequ}}{\PE\Deri_2}{\PE B_1}{\PE A_1}$,
  where $\PE A_1 =\PE A\sigma$ is the result of applying $\sigma$
  to~$\PE A$. To $\Deri_2$ we can apply a permutation argument,
  permuting all instances of $\acD$ up until they either reach the top
  of the derivation or an instance of $\me$ which separates the two
  atoms in the premise. More precisely, we consider the following inferene rule
  \begin{equation}
    \vlinf{\acDeq}{}{\Scons{a}}{S_0\cons{S_1\cons{a}\vlor S_2\cons{a}}}
  \end{equation}
  where $S_1\conhole\fequ\conhole\vlor E$ and
  $S_2\conhole\fequ\conhole\vlor F$ and $S\conhole\fequ
  S_0\cons{\conhole\vlor E\vlor F}$ for some formulas $E$ and $F$. The
  rule $\acDeq$ permutes over $\fequ$ and other instances of $\acDeq$,
  and over instances of $\me$ if they occur inside $S_0$ or $S_1$ or
  $S_2$. The only situation in which $\acDeq$ cannot be permuted up is
  the following:
  \begin{equation}
    \vlderivation{
      \vlin{\acDeq}{}{\Scons{R\cons{a}\vlan(C\vlor D)}}{
        \vlin{\me}{}{\Scons{(R_1\cons{a}\vlor R_2\cons{a})\vlan(C\vlor D)}}{
          \vlhy{\Scons{(R_1\cons{a}\vlan C)\vlor (R_2\cons{a}\vlor D)}}}}}
  \end{equation}
  Now we translate 
\end{proof}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
